---
title: "Reproducible Research - Week 4 Project 2"
author: "Joaquin Arango"
date: "12/28/2020"
output:
  pdf_document: default
  html_document: default
---
#NOAA Storm Database

For this project we will look at the NOAA Storm Database and answer questions about severe weather events and its impacts on health and ecomonics. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Synopsis 

Severe weather events can cause grave health and economic impacts across the US. Events that can results in injuries, damages or even death. The NOAA database keeps track of major serve events in the US.

##Questions: This data analysis must address the following questions:

* 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

* 2: Across the United States, which types of events have the greatest economic consequences?

## Data Processing and Retrieval


```{r}
#set the environment
library(knitr)
library(plyr)
library(ggplot2)
library(lattice)
library(data.table)
library(grid)
library(gridExtra)

File_data <- "repdata_data_StormData.csv.bz2"

# reading data
df <- read.csv(file = File_data, header=TRUE, sep=",")
```


## Select Useful data for Analysis once you read up on NOAA's Storm Data Documentation 
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf


```{r}
## variables selected
##Event Type (EVTYPE), Begin Date of Event (BGN_DATE), Event Human Fatalities (FATALITIES), Event Human Injuries (INJURIES), Event Estimated Property Damages (PROPDMG), Event Property Damage Decimal Precision (PROPDMGEXP), Event Estimated Crop Damages (CROPDMG), Event Property Damage Decimal Precision (CROPDMGEXP)
df <- df[ , c("EVTYPE", "BGN_DATE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
#changing date to POSIXct date
df$BGN_DATE <- as.POSIXct(df$BGN_DATE, format="%m/%d/%Y %H:%M:%S")
#take a look at current Data structure
str(df)
#look at list of events
head(unique(df$EVTYPE))
```
## Question 1 

Best way to answer this question is to look at fatalities and injuries as they are the most harmful with respect to health. 

Data Processing for Question 1:
```{r}
#add up fatalities by event type
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=df, sum)
injuries <- aggregate(INJURIES ~ EVTYPE, data=df, sum)
#arrange fatalities and injuries from highest to lowest by event type 
fatalities <- arrange(fatalities, desc(FATALITIES),EVTYPE)[1:10,]
injuries <- arrange(injuries, desc(INJURIES), EVTYPE)[1:10,]

#convert EVTYPE to factor so we can analyze
fatalities$EVTYPE <-factor(fatalities$EVTYPE, level = fatalities$EVTYPE)
injuries$EVTYPE <- factor(injuries$EVTYPE, levels=injuries$EVTYPE)
```

Question 1 Tables showing fatalities and injures by most common events.

```{r}
#to show table of fatalities
fatalities
```
```{r}
#to show table of injuries
injuries
```

Question 1 Plots 
```{r}
#plot fatalities
fatalitiesbyevent <- ggplot(fatalities, aes(x=EVTYPE, y=FATALITIES))+ geom_bar(stat = "identity", fill="red", width=NULL) +theme(axis.text.x = element_text(angle=90, hjust=1)) + xlab("Event Type") +ylab("Fatalities")

#plot injuries
injuriesbyevent <- ggplot(injuries, aes(x=EVTYPE, y=INJURIES))+ geom_bar(stat = "identity", fill="red", width=NULL) +theme(axis.text.x = element_text(angle=90, hjust=1)) + xlab("Event Type") +ylab("Injuries")

#plot both side by side
grid.arrange(fatalitiesbyevent, injuriesbyevent, ncol=2, nrow=1,
top = textGrob("Fatalities and Injuries from Top 10 Weather Related Events", gp=gpar(fontsize=14, font=3)))

```

## Question 2

Since question 2 asked for events with the greatest economic damage, we will use PROPDMGEXP and CROPDMGEXP, since they are in character format, we will need to convert them into numeric values.

```{r}
#turn from character into integer
numPROPDMG <- mapvalues(df$PROPDMGEXP,      c("K","M","","B","m","+","0","5","6","?","4","2","3","h","7","H","-","1","8"),c(1e3,1e6, 1, 1e9,1e6,1,1,1e5,1e6,1,1e4,1e2,1e3,  1,1e7,1e2,1,10,1e8))
numCROPDMG <- mapvalues(df$CROPDMGEXP, c("","M","K","m","B","?","0","k","2"), 
c( 1,1e6,1e3,1e6,1e9,1,1,1e3,1e2))
#Numeric Property and Crop Damage
df$TOTAL_PROPDMG <- as.numeric(numPROPDMG) * df$PROPDMG #property 
df$TOTAL_CROPDMG <- as.numeric(numCROPDMG) * df$CROPDMG #crop

#show columns
colnames(df)
```
```{r}
#lets create a total damage amount with property and crop
df$TOTALDMG <- df$TOTAL_PROPDMG + df$TOTAL_CROPDMG

#add up the total property and crop damage by event type
propdamage <- aggregate(TOTAL_PROPDMG ~ EVTYPE, data=df, sum)
cropdamage <- aggregate(TOTAL_CROPDMG ~ EVTYPE, data=df, sum)

#add total damages by event type
totaldamage <- aggregate(TOTALDMG ~ EVTYPE, data=df, sum)

#arrange the damages from highest to lowest for property and crop and total

propdamage <-arrange(propdamage, desc(propdamage$TOTAL_PROPDMG), EVTYPE)[1:10,]
cropdamage <- arrange(cropdamage, desc(cropdamage$TOTAL_CROPDMG),EVTYPE)[1:10,]
totaldamage <- arrange(totaldamage, desc(totaldamage$TOTALDMG), EVTYPE)[1:10,]

#change weather event type to factor variable

propdamage$EVTYPE <- factor(propdamage$EVTYPE, levels= propdamage$EVTYPE)
cropdamage$EVTYPE <- factor(cropdamage$EVTYPE, levels=cropdamage$EVTYPE)
totaldamage$EVTYPE <- factor(totaldamage$EVTYPE, levels=totaldamage$EVTYPE)

```

Question 2 Tables showing property and crop damage by most common events.

```{r}
#show table of Property Damage
propdamage
```

```{r}
#plot of Property Damage
propplotdamage <- ggplot(propdamage, aes(x=EVTYPE, y=TOTAL_PROPDMG)) + geom_bar(stat = "identity", fill="red") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("Event Type") +ylab("Property Damages in $")
#plot(propplotdamage) 
```

```{r}
#show table of Crop Damage
cropdamage
```


```{r}
#plot of Crop Damage
cropplotdamage <-ggplot(cropdamage, aes(x=EVTYPE, y=TOTAL_CROPDMG)) + geom_bar(stat = "identity", fill="red") + theme(axis.text.x = element_text(angle = 90, hjust =1)) + xlab("Event Type") +ylab("Crop Damages in $")
#plot(cropplotdamage) 
```

```{r}
#show table of Total Damage
totaldamage
```

```{r}
#plot of Total Damage
totalplotdamage <-ggplot(totaldamage, aes(x=EVTYPE, y=TOTALDMG)) + geom_bar(stat = "identity", fill="red") + theme(axis.text.x = element_text(angle = 90, hjust =1)) + xlab("Event Type") +ylab("Total Damages in $")
#plot(totalplotdamage) 

grid.arrange(propplotdamage, cropplotdamage, totalplotdamage, ncol=3, nrow=1,
     top = textGrob("Damages in $ - Property, Crop, & Total from Top 10 Weather Events ",gp=gpar(fontsize=12,font=2)))

```

# Results 

Question 1: Tornadoes caused the most fatalities, tornadoes caused the most significant harm out of all types of severe weather events.Heat and flash flooding are second and third leading causes of fatalities.

Tornado also cause more injuries. Thunderstorm related events (wind, lightning, flooding) combined are a significant injury contributor as well.

Question 2: 

Floods caused the most significant total damage for property and crop.

While property damage is most significant with flooding, Hurricanes/Typhoons, and Tornadoes, and Storm Surge are also strong contributors to property damage as well

Crop damage occurs with flooding, but drought events are the most significant cause of economic harm. For crops flooding, ice storms, and hail are also noted as main events to economic harm.